---
title: "Rapid assessment of Pacific Salmon spawner abundance to aide with COSEWIC prioritizaiton"
author: "Dylan Glaser (DFO)"
output: rmarkdown::github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, fig.path="output/plots/")
```
## Background
The purpose of this document is to import recent data on pacific salmon spawner abundances, and their designateable and conservation units (i.e. DUs and CUs). We plot abundances and examine trends, then explore various metrics to  prioritize populations for assessment. 
```{r pkgs, echo=FALSE}
library("tidyverse")
library("ggh4x") #hacks for making facets the same size 
```
### Load and wrangle data
We want to load the most recent data from a [GitHub repo](https://github.com/hertzPSF/COSEWIC-compilation) managed by Eric Hertz at the Pacific Salmon Foundation (PSF). We download the repo, import data we want from it, and tidy it up. 
```{r read in data}
download.file(url="https://github.com/hertzPSF/COSEWIC-compilation/archive/master.zip",
              destfile="compilation/hertzPSF-compilation.zip")

unzip(zipfile="compilation/hertzPSF-compilation.zip", exdir="compilation")

erics_output <- list.files("compilation/COSEWIC-compilation-main/output")

raw_sp_data <- read.csv(paste0("compilation/COSEWIC-compilation-main/Output/",
                           erics_output[grep("CU_Spawner_Abund", erics_output)]), 
                        header=TRUE) %>%
         mutate(logSp=log(Spawner.Abundance))

cu_metadata <- read.csv(paste0("compilation/COSEWIC-compilation-main/Output/",
                               erics_output[grep("CU_Metadata", erics_output)]), 
                        header=TRUE, na.strings=c("", "NA")) %>%
  filter(!is.na(gen_length)) %>%
  rename(CUID=cuid)

#join, dump pops with NA DUs, and create a du_cu key
sp_data <- left_join(raw_sp_data, select(cu_metadata, CUID, du_number, gen_length), 
                     by="CUID") %>%
  drop_na(du_number, Spawner.Abundance) %>%
  mutate(du_cu=paste(du_number, CUID)) %>%
  relocate(du_cu, 1) %>%
  arrange(du_cu, Year)
```
#### Get the last 3 generations of spawner abundances
We need to use the metadata to get generation times for each population, then use that to calculate a time span that includes the last 3 generations, as per IUCN guidelines. 

```{r}
last3_gens <- sp_data %>%
  group_by(du_cu) %>%
  mutate(three_gens = gen_length*3) %>%
  filter(Year > (last(Year) - three_gens), Year <= last(Year))
```

```{r, echo = FALSE}
rm(raw_sp_data, erics_output, cu_metadata) #remove some helper objects
```
## Plot spawner abundance
We'll make 2 plots for each species and region: raw spawner abundance and ln-transformed spawner abundance with a linear regression passing through all data points (black dashed line) and the last 3 generations (red). **Note that that the y axis varies for ease of viewing**.    
This code is cumbersome in order to make it robust, so we've omitted it from this document. See the [.Rmd document](https://github.com/Pacific-salmon-assess/COSEWIC-prioritize/blob/main/prioritize.Rmd) for details.   

```{r abundance plots, echo = FALSE}
max_facets <- 20 #how many facets (i.e. panes) per plot?
rows <- 4 #how many rows/cols? note needs to be multiple of facets. 
cols <- 5

for(i in unique(sp_data$Species)){
  sp_data_species <- filter(sp_data, Species==i)
  for(j in unique(sp_data_species$Region)){
    sub_data <- filter(sp_data_species, Region==j) %>%
      complete(Year, Species, Region, du_cu) #explicitly include NA so line draws right
    sub_last3 <- filter(last3_gens, Species==i, Region==j)
    n_pops <- length(unique(sub_data$du_cu))
    if( n_pops>max_facets){
      pages <- ceiling(n_pops/max_facets)
    for(k in 1:pages){
      sub_data_2 <- filter(sub_data, du_cu %in% 
                               unique(sub_data$du_cu)[((k-1)*max_facets)+1:(max_facets*k)])
      
      sub_last3_2 <- filter(sub_last3, du_cu %in% 
                                unique(sub_last3$du_cu)[((k-1)*max_facets)+1:(max_facets*k)])
      
      p <- ggplot(data = sub_data_2, aes(x=Year, y=Spawner.Abundance/1000)) +
        geom_line(color="grey") +
        geom_point() +
        ggh4x::facet_wrap2(vars(du_cu), nrow=rows, ncol=cols, trim_blank=FALSE,
                           scales="free_y") +
        labs(x="Year", y="spawners (thousands)", 
             title = paste0(j," ",i," spawner abundance ","(",k," of ",pages,")")) +
        theme_bw() +
        theme(strip.background=element_blank(), axis.text.x=element_text(angle = 45))
      print(p)
      
      p_log <- ggplot(data=sub_data_2, aes(x=Year, y=logSp)) +
        geom_line(color="grey") +
        geom_point() +
        stat_smooth(method="lm", formula=y~x, color="black", lty="dashed", se=FALSE) +
        stat_smooth(data = sub_last3_2, formula=y~x, method="lm", color="red", se=FALSE, 
                    na.rm=TRUE) +
        ggh4x::facet_wrap2(vars(du_cu), nrow=rows, ncol=cols, trim_blank=FALSE,
                           scales="free_y") +
        labs(x="Year", y="ln(spawners)", 
             title=paste0("ln ",j," ",i," spawner abundance ","(",k," of ",pages,")")) +
        theme_bw() +
        theme(strip.background=element_blank(), axis.text.x=element_text(angle=45))
      print(p_log)
    }
  }else{
    p <- ggplot(data=sub_data, aes(x=Year, y=Spawner.Abundance/1000)) +
      geom_line(color="grey") +
      geom_point() +
      ggh4x::facet_wrap2(vars(du_cu), nrow=rows, ncol=cols, trim_blank=FALSE, 
                         scales="free_y") +
      labs(x="Year", y="spawners (thousands)", title=paste(j, i, "spawner abundance")) +
      theme_bw() +
      theme(strip.background=element_blank(), axis.text.x=element_text(angle = 45))
    print(p)
    
    p_log <- ggplot(data=sub_data, aes(x=Year, y=logSp)) +
      geom_line(color="grey") +
      geom_point() +
      stat_smooth(formula=y~x, method="lm", color="black", lty="dashed", se=FALSE) +
      stat_smooth(data=sub_last3, formula=y~x, method="lm", color="red", se=FALSE, 
                  na.rm=TRUE) +
      ggh4x::facet_wrap2(vars(du_cu), nrow=rows, ncol=cols, trim_blank=FALSE, 
                         scales="free_y") +
      labs(x="Year", y="ln(spawners)", title=paste(j, i, "ln(spawner abundance)")) +
      theme_bw() +
      theme(strip.background=element_blank(), axis.text.x=element_text(angle = 45))
    print(p_log)
  }
 }
}
```
```{r, echo=FALSE}
#clean environment for my own sanity
rm(p, p_log, sp_data_species, sub_data, sub_data_2, sub_last3, sub_last3_2)
```
## Linear models of population decline
#### Which populations are data deficient?
Define a threshold of acceptable data coverage then get a list of DUs that don't meet this threshold and filter them out of data that will be used in the population decline models.  
First we'll do this for the recent data.
```{r}
deficient_recent_perc <- 0.7

deficient_recent <- last3_gens %>%
  group_by(du_cu) %>%
  summarise(n = n(), #years with data  
            n_proper = mean(gen_length)*3) %>% #years for 3 gens of data if fully sampled
  mutate(perc_na = n/n_proper) %>%
  filter(perc_na < deficient_recent_perc) %>%
  pull(du_cu)

last3_gens_filtered <- filter(last3_gens, !(du_cu %in% deficient_recent))
```
Then we'll use similar logic but account for the whole time series using a more forgiving 
cutoff and the first year of sampling. 
```{r}
deficient_historical_perc <- 0.6

deficient_historical <- sp_data %>%
  group_by(du_cu) %>%
  summarise(n = n(),
            n_proper = last(Year) - first(Year) + 1) %>%
  mutate(perc_na = n/n_proper) %>%
  filter(perc_na < deficient_historical_perc) %>%
  pull(du_cu)

sp_data_filtered <- filter(sp_data, !(du_cu %in% deficient_historical))
```
## Population decline  
### Basic linear models  
Now we'll run basic linear models on ln(spawner abundances) for the whole time series and
for the last 3 generations. We need to add some if statements in this loop to account for
which du_cus made it through the data filters above.   
```{r}
summary_table <- NULL

for(i in unique(sp_data_filtered$du_cu)){
  sub_raw <- filter(sp_data, du_cu==i)
  sub_data <- filter(sp_data_filtered, du_cu==i)
  sub_last3 <- filter(last3_gens_filtered, du_cu==i)
  if(nrow(sub_data)==0){
    slope_all <- NA
  } else{
    if(nrow(sub_last3)==0){
      slope_recent <- NA
    } else{
      model_all <- lm(sub_data$logSp~sub_data$Year)
      model_recent <- lm(sub_last3$logSp~sub_last3$Year)
      slope_all <- round(as.numeric((exp(model_all$coefficients[2]*nrow(sub_data))-1)*100))
      slope_recent <- round(as.numeric((exp(model_recent$coefficients[2]*nrow(sub_last3))-1)*100))
    }
  }
  recent_abundance <- last(sub_raw$Spawner.Abundance)
  last_year_monitored <- last(sub_raw$Year)
  
  output <- data.frame(du_cu=i, slope_recent, slope_all, recent_abundance, last_year_monitored)
  
  summary_table <- bind_rows(output, summary_table)
}

head(summary_table)
```

```{r, echo = FALSE}
rm(list=ls()[grep("sub|model|def|slope", ls())]) #yeet garbage
```

### Bayesian estimates with the MetricsCOSEWIC package
First some housekeeping. You can use devtools to download the [MetricsCOSEWIC package](https://github.com/SOLV-Code/MetricsCOSEWIC) from GitHub. We'll wrangle the data into the format this package wants as well. We'll feed it the filtered last 3 generations.   
**note these are preliminary as the MetricsCOSEWIC package is still in development**  
*Still need to make sure I did this right. The README says the first argument to calcPercChangeMCMC() is a "vector with numeric values" so I'll assume that is spawner abundance with NAs omitted?*  
```{r Bayesian lms, results = "hide"}
library(MetricsCOSEWIC)
slope_posterior <- NULL
jags_summary <- NULL

for(i in unique(last3_gens_filtered$du_cu)){
  sub_data <- filter(last3_gens_filtered, du_cu == i)
  jags_mod <- calcPercChangeMCMC(sub_data$Spawner.Abundance, method="jags", out.type="long") 
  
  summary <- data.frame(jags_mod$summary) %>%
    slice(1:3) %>%
    select(mean, sd, X2.5., X50., X97.5., n.eff, Rhat) %>%
    rename(perc_2.5 = X2.5., 
           perc_97.5 = X97.5.,
           median = X50.) %>%
    mutate(du_cu = i,
           region = sub_data$Species[1],
           species = sub_data$Region[1]) %>%
    relocate(du_cu, region, species, .before = mean) 
  
  rownames(summary) <- NULL #rownames to columns - probably a better way
  summary$var <- c("intercept", "slope", "sigma")

  jags_summary <- bind_rows(summary, jags_summary)

  slope_posterior <- bind_rows(data.frame(du_cu =i, region = unique(sub_data$Region),
                                          species = unique(sub_data$Species),
                                          draw = jags_mod$samples$slope), slope_posterior)
}

jags_summary <- relocate(jags_summary, var, .before = mean)
```
We'll also create an object that summarizes how many draws are in each category. This will be a go-to summary table for COSEWIC thresholds for population decline based on [category A2](https://www.canada.ca/en/environment-climate-change/services/species-risk-act-accord-funding/listing-process/quantitative-criteria-guidelines-status-table-2.html), where 50 and 30 percent reductions correspond to endangered and threatened status, respectively. we'll use 30, 50, and 70 % reductions as thresholds for least concern, threatened, and endangered for now.  
**these categories could be revisited**
```{r categories of decline}
perc_category <- slope_posterior %>%
  group_by(species, region, du_cu) %>%
  summarise(not_at_risk = round(sum(draw>(-30))/n()*100, 2),
            least_concern = round(sum((-30)>draw & draw>(-50))/n()*100, 2), 
            threatened = round(sum(draw<(-50) & draw>(-70))/n()*100, 2), 
            endangered = round(sum(draw<(-70))/n()*100, 2))
head(perc_category)
```
And we can arrange this table to see which populations are in the roughest shape.  
```{r category sort}

```

### Plot slope posterior distributions  
These distributions show the probability (i.e. each markov chain monte carlo draw) that the slope of the last 3 generations is in of the categories listed above. The distribution is plotted over the decline categories specified above. It's essentially a visualization of the perc_category dataframe.  
The amount of the distribution in the green region means the population is experiencing less than 30% declines, the red region says the population is declining by 70% or more, and yellow is intermediate.  
```{r slope posteriors, echo = FALSE}
#helper function to center limits
symmetric_limits <- function(x){
    max <- max(abs(x))
    c(-max, max)
}

for(i in unique(slope_posterior$species)){
  post_species <- filter(slope_posterior, species == i)
  for(j in unique(post_species$region)){
    sub_data <- filter(post_species, region == j)
    if(length(unique(sub_data$du_cu)) > max_facets){
      pages <- ceiling(length(unique(sub_data$du_cu))/max_facets)
    for(k in 1:pages){
      sub_data_2 <- filter(sub_data, du_cu %in% 
                               unique(sub_data$du_cu)[((k-1)*max_facets)+1:(max_facets*k)])
      p <- ggplot(data = sub_data_2, aes(draw)) +
        geom_density(fill = "gray") + 
        scale_x_continuous(limits = symmetric_limits) +
        geom_rect(aes(xmin=-30, xmax=Inf, ymin=0, ymax= Inf), fill="green", alpha=0.002) +
        geom_rect(aes(xmin=-50 , xmax=-30, ymax=Inf, ymin=-Inf), fill="yellow", alpha=0.002) +
        geom_rect(aes(xmin=-70, xmax=-50, ymax=Inf, ymin=-Inf), fill="orange", alpha=0.002) +
        geom_rect(aes(xmin=-Inf, xmax=-70, ymax=Inf, ymin=-Inf), fill="red", alpha=0.002) +
        theme_classic() +
        ggh4x::facet_wrap2(vars(du_cu), nrow=rows, ncol=cols, trim_blank=FALSE, 
                           scales="free") +
        labs(title = paste(j, i, "slope posterior", "(", k, " of ", pages, ")"), 
         x = "MCMC draws") +
        guides(x = "none", y = "none")
      print(p)
    }
  }else{
      p <- ggplot(data = sub_data, aes(draw)) +
        geom_density(fill = "gray") + 
        scale_x_continuous(limits = symmetric_limits) +
        geom_rect(aes(xmin=-30, xmax=Inf, ymin=0, ymax= Inf), fill="green", alpha=0.002) +
        geom_rect(aes(xmin=-50 , xmax=-30, ymax=Inf, ymin=-Inf), fill="yellow", alpha=0.002) +
        geom_rect(aes(xmin=-70, xmax=-50, ymax=Inf, ymin=-Inf), fill="orange", alpha=0.002) +
        geom_rect(aes(xmin=-Inf, xmax=-70, ymax=Inf, ymin=-Inf), fill="red", alpha=0.002) +
        theme_classic() +
        ggh4x::facet_wrap2(vars(du_cu), nrow=rows, ncol=cols, trim_blank=FALSE, 
                           scales="free") +
        labs(title = paste(j, i, "slope posterior"), x = "MCMC draws") +
        guides(x = "none", y = "none")
      print(p)
  }
 }
}
```